Moment Constraints Make Learning Latent-Variable Models Easier

### 2014-01-31 doc/icml2014/abstract.tex
Added words and random mumbo-jumbo
Updated directed section
Updated draft

\begin{abstract} 
\end{abstract}

\paragraph{Errata}
\begin{enumerate}
  \item The likelihoods as defined are concave, not convex.
  \item Define parameters $\theta$ more uniformly.
\end{enumerate}

>>>>

\begin{abstract}

Parameter estimation in latent variable models is prone to local
  optima due to the non-convexity of the likelihood. 
In this paper, we extend recent advances in the method of moments 
  to provide consistent parameter estimates for a general class of
  directed and undirected graphical models, thus avoiding the issue of
  local optima.
In particular, our algorithm applies to any graphical model wherein each
  hidden variable has at least three conditionally independent observed
  variables.
This family includes loopy graphs with high treewidth.
We use the method of moments to tie each hidden to some observed
  variables, and show that the rest of the problem can be cast as convex
  optimization of the likelihood.
This approach gracefully extends to models outside this family by
  providing partial information via posterior regulraization.
\end{abstract}

\paragraph{Last updated:} \today, \currenttime

\paragraph{Errata / Incomplete}
\begin{enumerate}
  \item Use \texttt{\textbackslash left(}.
\end{enumerate}

### 2014-01-31 doc/icml2014/background.tex
Updated draft

Let $\sG$ be a discrete graphical model with variables $V$, 
  let the subset $X \subseteq V$ be observed and 
  let $H = V \setminus O$ be hidden.
Let $\Pa(v) \subset V$ be the parents of a variable $v$ in a directed
  model, and $\sN(v) \subset V$ be the neighbours of $v$ in an undirected
  model.
We will use $x_1, x_2, \cdots, x_N \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For ease of exposition, we assume that $x_i \in \{1, \cdots, d\}$ for
  all $i$, and that $h_i \in \{1, \cdots, k\}$ for all i.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    \sL &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh} p_\theta(\vx,\vh)
  \end{align*}
  which is statistically efficient but computationally intractable
  because we must marginalize over latent variables $z$.
In practice, one uses gradient-based optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.

@@

We use $\otimes$ to denote the tensor product, e.g if $u \in \Re^k,
  v \in \Re^d$, then $u \otimes v \in \Re^{k \times d}$.
For a $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $x_1, \cdots, x_\ell \in \Re^{d},$, we define 
\begin{align*}
  T(x_1, \cdots, x_\ell) 
      &= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
            T_{i_1 i_2 \ldots i_l} ({x_1})_{i_1} \cdots ({x_\ell})_{i_\ell}.
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)_{j_1, \ldots, j_l} 
      &= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
            T_{i_1 i_2 \ldots i_l} ({M_1})_{i_1j_1} \cdots ({M_\ell})_{i_\ell j_\ell}.
\end{align*}

@@

Recently, \citet{anandkumar12moments} presented a consistent algorithm
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models, that sidesteps the local optima problem altogether. 
In the model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a latent variable $h$.  
The parameters of this model are the mixture probabilities $\pi \eqdef
  \Pr(h) \in \Delta_{k-1}$ and the conditional moments $O^{(i)} \eqdef
  \Pr(x_i \given h) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters;
\begin{align*}
  \mO_{ij} &= \sum_h \pi_h {O\oft i}^T_h \otimes {O\oft j}^T_h & \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h \pi_h {O\oft 1}^T_h \otimes {O\oft 2}^T_h \otimes  {O\oft 3}^T_h.
\end{align*}

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts consistent estimates of the observed moments 
  into consistent estimates of the parameters, $\pi, O\oft{1},
  O\oft{2}, O\oft{3}$ (up to permutation of the columns), provided
  the following assumption holds:

\begin{assumption}
\label{asm:full-rank}
Assume the conditional moments $O\oft{1}, O\oft{2}, O\oft{3}$ have full column rank
  $k$, and that $\pi \succ 0$.
\end{assumption}

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
  of the whitened $\mO_{123}$; unwhitening yields the columns of $\mO_{3}$.
$\mO_1$ and $\mO_2$ can then be recovered with some algebra.

>>>>

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
Let $\Pa(a) \subset X \union H$ be the parents of a variable $a$ in a directed
  model, and $\sN(a) \subset X \union H$ be the neighbours of $a$ in an
  undirected model.
We will use $x_1, x_2, \cdots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For ease of exposition, we assume that $x_i \in \{1, \cdots, d\}$ for
  all $i$, and that $h_i \in \{1, \cdots, k\}$ for all $i$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef
  \underbrace{d \times d \cdots \times d}_{L ~\textrm{times}}$ and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef
  \underbrace{k \times k \cdots \times k}_{M ~\textrm{times}}$.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    \sL &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh)
  \end{align*}
  which is statistically efficient but computationally intractable
  because we must marginalize over latent variables $\vh$.
In practice, one uses gradient-based optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\vec
  i = (i_1, \cdots, i_l)$, $T[\vec i] = T_{i_1 \ldots i_l}$.
Similarly, for a set $\sC$ of size $k$ and vector $\vec x \in \Re^d$, we
  will use the notation $\vec x_\sC$ to get a vector $\Re^k$ which selects
  the appropriate elements of $\vec x$

We use $\otimes$ to denote the tensor product, e.g if $u \in \Re^k,
  v \in \Re^d$, then $u \otimes v \in \Re^{k \times d}$.
For a $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $x_1, \cdots, x_\ell \in \Re^{d},$, we define 
\begin{align*}
  T(x_1, \cdots, x_\ell) 
      &= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
            T[i_1, i_2, \ldots, i_l] x_1[i_1] \cdots x_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      &= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell],
\end{align*}
where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

@@

Recently, \citet{anandkumar12moments} presented a consistent algorithm
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models, that sidesteps the local optima problem altogether. 
In the model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a latent variable $h$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h) \in \Delta_{k-1}$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters;
\begin{align*}
  \mO_{ij} &= \sum_h {\pi \oft 1}[h] {\mOppt{i}{1}}[h] \otimes {\mOppt{j}{1}}[h] \quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{i}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{2}{1}}[h].
\end{align*}

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts consistent estimates of the observed moments 
  into consistent estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
  the following assumption holds:

\begin{assumption}
\label{asm:full-rank}
Assume the conditional moments $\mOpp{1}{1}, \mOpp{2}{1}, \mOpp{3}{1}$ have full column rank
  $k$, and that $\pi\oft 1 \succ 0$.
\end{assumption}

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
  $\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with some algebra.
